Reproducible Research: Peer Assessment 1
========================================================
This analysis makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps
taken in 5 minute intervals each day.

```{r setoptions, echo = FALSE}
opts_chunk$set(echo = TRUE, warning = FALSE)
wd<- "C:/Users/jsardinha/Documents/GitHub/RepData_PeerAssessment1"
setwd(wd)
```

## Loading and preprocessing the data
This repository contains the dataset, hence there is not need to download 
it seperatly. The following code block downloads the activity.csv data into R
```{r dataload}
library(plyr)
library(ggplot2)
library(xtable)

zipFile <- "activity.zip"
unzip(zipFile)
activityData <- read.csv("activity.csv")
```

Let's analyze the imported activityData fata.frame
```{r dataAnalysis}
print(str(activityData))
```

After loading the dataset, it is apparent that the $date variable is of class
"factor". Hence, the following steps will convert it to type "date".
Note: Missing values coded as 'NA' are not handelled here and will be handelled in subsequent steps.
```{r dateconvert}
activityData$date <-as.Date(activityData$date)
print(str(activityData))
```

Now we have a clean data set to work with.

## What is mean total number of steps taken per day?

Plotting the total steps taken each day
```{r sumSteps}
activityData2 <-ddply(activityData, ~date , summarize, 
                      sum_steps = sum(steps),
                      mean_steps = mean(steps),
                      median_steps = median(steps)) 
g <- ggplot(activityData2, aes(x = date, y = sum_steps))
g <- g + geom_bar(stat="identity")
g <- g + xlab("Date") + ylab("Total number of steps taken") + labs(title = "Total number of steps taken per day")
print(g)
```

Displaying the mean and median steps taken each day
```{r }
print((activityData2))
```


## What is the average daily activity pattern?

```{r}
activityData3 <-ddply(activityData, ~interval , summarize, 
                     mean_steps = mean(steps, na.rm = TRUE))
yMax<- max(activityData3[,2])
xMax <- activityData3[which.max(activityData3[,2]),1]
g <- ggplot(activityData3, aes(x = interval, y = mean_steps))
g <- g + geom_line()
labelValue <- paste("Max value at (", as.character(xMax), ",", yMax, ")", sep = "")
g <- g + annotate("text", x=xMax, y=yMax, label=labelValue)
g <- g + xlab("Interval") + ylab("Average number of steps taken") + labs(title = "Average number of steps taken per interval on all days")
print(g)
```


## Imputing missing values

### Strategy for dealing with missing values
Finding the total number of rows with missing values (coded as NA).
```{r}
summary(activityData)
```
from this we see that NA's only exist in the $steps variable and that there are 2304 of them.




## Are there differences in activity patterns between weekdays and weekends?
